---
title: "Vernal pool vegetation"
subtitle: "Week 4"
categories: [library(), tidyverse, here, naniar, patchwork, read_csv(), filter(), mutate(), unite(), str_detect(), left_join(), group_by(), summarize(), ungroup(), ggplot(), geom_point(), geom_line(), gg_miss_var(), gg_miss_upset, quarto]
---

# Summary

## Guiding questions

What are the temporal trends in Coyote-thistle (Eryngium vaseyi) percent cover across years?

What differences are there in Coyote-thistle percent cover across vernal pools?

## Data science tools

To answer these questions, we are going to:  

- review reading in packages
  - `library()`
- use familiar packages
  - `tidyverse`
  - `here`
- introduce new packages
  - `naniar` to visualize missing data
  - `patchwork` to arrange multipanel figures
- review reading in data
  - `read_csv()`
- review cleaning and wrangling functions/operators
  - `filter()`
  - `mutate()`
  - `left_join()` 
- review summarizing functions 
  - `group_by()`, `summarize()`, and `ungroup()`
- review visualizations using `ggplot()`, including:
  - `geom_point()` and `geom_line()`
- introduce a new visualization tool:
  - `gg_miss_var()` and `gg_miss_upset()` to visualize missing data
- introduce new wrangling functions:
  - `unite()` from the `tidyr` package (within `tidyverse`) to create new columns with values separated by a character
  - `str_detect()` from the `stringr` package (within `tidyverse`) to detect any values in a character string


# In-class code

## 1. Set up

### Packages

```{r}
#| label: packages
#| message: false

# general use
library(tidyverse)

# file organization
library(here)

# data visualization
library(naniar)
library(patchwork)
```

### Data

```{r}
#| label: data

veg <- read_csv(here("data", "veg.csv"))

metadata <- read_csv(here("data", "vp_veg_metadata.csv"))
```

## 2. Cleaning

### Exploring missingness prior to cleaning

```{r}
#| label: missingness-plots

gg_miss_var(veg)
gg_miss_upset(veg)
```

**Which variables tend to have missing values?**

[insert note here]

### Implicit NAs

:::{.column-margin}
edited 2026-04-22: adding:

```{r}
#| eval: false

  filter(str_detect(transect_name, "VP") &
           site == "ncos") 
```

to cleaning code in implicit NA section _and_ explicit 0 section to filter for NCOS sites in addition to any transect names with a character string of `"VP"`.
:::

```{r}
#| label: implicitNAs

# create a new object from veg
pools_implicitNA <- veg |> 
  # filter all occurrences where transect_name contains VP
  # and only observations at NCOS
  filter(str_detect(transect_name, "VP") &
           site == "ncos") |> 
  # group by year, pool, and species
  group_by(year, transect_name, psoc) |>
  # calculate mean percent cover
  summarize(mean_pc = mean(percent_cover, na.rm = TRUE)) |> 
  # ungroup data frame
  ungroup() |> 
  # create a new column called year_pool from year and transect name
  unite("year_pool", year, transect_name, remove = TRUE) |>
  # join with metadata by year_pool
  left_join(metadata, by = "year_pool")
```

### Explicit 0s

```{r}
#| label: explicit0s

# create a new object from veg
pools_explicit0 <- veg |> 
  # filter all occurrences where transect_name contains VP
  # and only observations at NCOS
  filter(str_detect(transect_name, "VP") &
           site == "ncos") |> 
  # complete all possible combinations of year, pool, psoc and distance
  complete(year, transect_name, psoc, transect_distance,
           # fill values of percent cover with 0
           fill = list(percent_cover = 0)
           ) |>
  # group by year, pool, and psoc
  group_by(year, transect_name, psoc) |>
  # calculate total percent cover across all quads
  summarize(sum_pc = sum(percent_cover, na.rm = TRUE)) |>
  # ungroup the data frame
  ungroup() |>
  # create a new column called year_pool from year and transect name
  unite("year_pool", year, transect_name, remove = TRUE) |> 
  # join with metadata data frame
  left_join(metadata, by = "year_pool") |> 
  # calculate mean percent cover from total percent cover/number of quadrats
  mutate(mean_pc = sum_pc/num_quad)
```

### Exploring missingness after summarizing

```{r}
#| label: summarized-missingness-plots

gg_miss_var(pools_implicitNA)

gg_miss_var(pools_explicit0)
```

No missing values in either data frame.

## 3. Filtering further

Filter to only include _Eryngium vaseyi_.

```{r}
#| label: filtering-coyotethistle

# creating new ojbect from implicit NA data frame
filtered_implicitNA <- pools_implicitNA |> 
  # filter to only include 
  filter(psoc == "Eryngium_vaseyi")

filtered_explicit0 <- pools_explicit0 |> 
  filter(psoc == "Eryngium_vaseyi")
```

## 4. Visualizing

Create a plot depicting mean percent cover through time using the implicit NA data frame.

Color geometries by vernal pool.

```{r}
#| label: implicitNA-plot

# creating a new object to store plot
implicitNA_plot <- ggplot(data = filtered_implicitNA, 
                          # x-axis: year
                          # y-axis: mean percent cover
                          # color geometries by pool
                          aes(x = year,
                              y = mean_pc,
                              color = transect_name)) +
  # first layer: points to represent mean percent cover
  geom_point() +
  # second layer: lines to connect points
  geom_line() +
  # add title to describe data frame
  labs(title = "no complete")

# display plot
implicitNA_plot
```

Create a plot depicting mean percent cover through time using the explicit 0 data frame.

Color geometries by vernal pool.

```{r}
#| label: explicit0-plot

# creating a new object to store plot
explicit0_plot <- ggplot(data = filtered_explicit0, 
                          # x-axis: year
                          # y-axis: mean percent cover
                          # color geometries by pool
                          aes(x = year,
                              y = mean_pc,
                              color = transect_name)) +
  # first layer: points to represent mean percent cover
  geom_point() +
  # second layer: lines to connect points
  geom_line() +
  # add title to describe data frame
  labs(title = "complete (understanding missing values are 0)")

# display plot
explicit0_plot
```

Put both of those plots together using `patchwork`.

```{r}
#| label: multipanel
#| fig-height: 10
#| fig-width: 8

implicitNA_plot / explicit0_plot
```

**How do these plots differ?**

[insert note here]

**How does Coyote-thistle percent cover change through time? Is this different across pools?**

[insert note here]
