---
title: "Bird species at NCOS"
subtitle: "Week 2"
categories: [library(), tidyverse, ggplot colors, filter(), mutate(), select(), group_by(), summarize(), ungroup(), ggplot(), geom_bar(), facet_wrap(), reorder(), scale_color_manual(), distinct(), nrow(), colnames(), case_when(), count(), arrange(), n_distinct(), sum(), fct_relevel(), '%in%', quarto]
---

[Download zip here](https://github.com/ENVS-193DD/ENVS-193DD.github.io/blob/main/data/week-02.zip)

# Summary

## Guiding questions

Across four main bird groups, what are the:  

- species richness trends across seasons and through time?
- individual count trends across season and through time?
- individual count differences across seasons within a group?

## Data science tools

To answer these questions, we are going to:  

- review reading in packages
  - `library()`
- use familiar packages
  - `tidyverse`
- introduce new packages
  - `NatParksPalettes` to customize plot colors
- review reading in data
  - `read_csv()`
- review cleaning and wrangling functions from last week
  - `filter()`
  - `mutate()`
  - `select()`
- review summarizing functions from last week
  - `group_by()`, `summarize()`, and `ungroup()`
- review visualizations using `ggplot()`
- introduce a new visualization tools
  - `geom_bar()`
  - `facet_wrap()`
- make visualizations easier to intepret using `reorder()`
- customize colors in visualizations using
  - `scale_color_manual()`
- introduce new summarizing/exploration functions
  - `distinct()`
  - `nrow()`
  - `colnames()`
  - `case_when()`
  - `count()`
  - `arrange()`
  - `n_distinct()`
  - `sum()`
- introduce extracting values from dataframes using square brackets (`[]`) and references to rows and columns
- introduce a new wrangling function
  - `fct_relevel()`
- introduce a new operator: `%in%`

Additionally, we are going to learn new ways of working with code in a Quarto document (file suffix `.qmd`). Quarto allows people to write code along with prose.  

In class, we will discuss how to:  

- insert code chunks to run code
- render a document to turn a .qmd file into a .pdf
- set code chunk options to make sure the output is readable in a final rendered document
  - hide messages
  - hide warnings
  - adjust figure sizes
- use headers and code chunk labels to organize documents

# In-class code

## Set up

### Packages

```{r}
#| label: packages

# cleaning packages
library(tidyverse)

# visualization packages
library(NatParksPalettes)
library(scales)
```

### Data

```{r}
#| label: data for rendering
#| echo: false

# reading in data
birds <- read_csv(here::here("data", "birds.csv"))
```

```{r}
#| label: data
#| eval: false

# reading in data
birds <- read_csv("birds.csv")
```

## Look at the data

Look at the data by:

-   clicking on the object in the Environment (or running `View(birds)` in the Console)
-   running `class(birds)` in the Console to understand what type of object it is
    - this is a data frame
-   running `str(birds)` in the Console to look at the structure
    - there are 6 columns and 15,377 rows in this data frame

Another useful function to look at the data is `colnames()`, which gives you a list of column names in the data frame.  

Run `colnames(birds)` in the Console.  

Make a list of the variables in this data set, what they contain, and what they represent.

Columns: 

- `observation_date`: date of observation, which represents the date that individuals/species were seen
- `season`: character, which represents the season of observation
- `species`: character, which represents bird species
- `count`: numeric, species counts for a given observation
- `repeat_observation`: character, whether or not the observation was a repeated sighting
- `e_bird_group`: character, which represents group of birds according to eBird

## Figuring out unique eBird groups

We'll be working with four specific types of birds this week.  

The information about those groups are in the `e_bird_group` column of the data frame.

However, there are a lot of rows in the data frame, so it's hard to see what the eBird groups are. 

Use `distinct()` to figure out what the eBird groups are.

```{r}
#| label: distinct groups

# creating new data frame from birds
ebird_groups <- birds |> 
  # extracting unique values for e_bird_group column
  distinct(e_bird_group)
```

How many groups are there in this dataset?

```{r}
#| label: nrow to find number of groups

# counting number of eBird groups
nrow(ebird_groups)

```

There are 39 groups in this dataset.

## Wrangling the data

### Pseudocode

Write pseudocode to create a data frame called `birds_wy2025` that:

- only includes the 2025 water year,
- excludes repeat observations, and 
- includes observations from the following eBird groups: Waterfowl; New World Sparrows; Herons, Ibis, and Allies; Shorebirds

**Pseudocode:**

- create object called `birds_wy2025` from `birds`
- filter date to keep observations after Sept 30 2024 and before Oct 1 2025 
- filter repeat_observations to only include "No" rows
- filter ebird_group to only include Waterfowl; New World Sparrows; Herons, Ibis, and Allies; Shorebirds
  
### Running code

We'll add a new function at the end of this code to insert the lowest taxonomic group for each eBird group name.  

We'll use `mutate()` (a function you already know) along with `case_when()`, which is a new function, to replace values in a column.

```{r}
#| label: cleaning and wrangling birds

# creating new object from birds
birds_wy2025 <- birds |> 
  # filter to include:
  # observation_dates in the 2025 water year
  # observations that are not repeats
  # eBird groups in the 4 groups of interest
  filter(observation_date > "2024-09-30" & observation_date < "2025-10-01",
         repeat_observation == "No",
         e_bird_group %in% c("Waterfowl",
                             "New World Sparrows",
                             "Herons, Ibis, and Allies",
                             "Shorebirds")) |> 
  # create a new column called lowest group
  mutate(lowest_group = case_when(
    # lowest group for sparrows is family
    e_bird_group == "New World Sparrows" ~ "Passerellidae",
    # lowest group for herons is family
    e_bird_group == "Herons, Ibis, and Allies" ~ "Ardeidae",
    # lowest group for shorebirds is order
    e_bird_group == "Shorebirds" ~ "Charadriiformes",
    # lowest group for waterfowl is family
    e_bird_group == "Waterfowl" ~ "Anatidae"
  ))
```

### Figure out the species in each group.

You can combine wrangling functions to get subsets of information from a dataset.

At the end of this chunk of code, you should end up with a dataframe with the all the species in all eBird groups.

```{r}
#| label: finding species in a group

# create a new object called group_species from birds_wy2025
group_species <- birds_wy2025 |> 
  # group data frame by ebird group
  group_by(e_bird_group) |> 
  # find the unique species in each group
  distinct(species) |> 
  # arrange by group
  arrange(e_bird_group)
```

How many species are in each group? We can use the `count()` function to figure it out.

```{r}
#| label: finding number of species in a group

# create new
group_species_count <- group_species |> 
  count()

group_species_count
```

Note that we ran that code without `ungroup()` in the `group_species` code. What happens if you add `ungroup()` to that code and rerun the next chunk?  

Adding `ungroup()` leads to one number that represents all the species across all four eBird groups (37 total).

What if you wanted to extract a single number from the dataframe?  

You can subset data frames using square brackets `[ ]`. The syntax to extract a single value from a dataframe is as follows:  

`dataframe[row position, column position]`

So for example, to extract the value representing the total number of Shorebird species from `group_species_count` you would run something like:

```{r}
#| label: finding value for total number of shorebirds

group_species_count[3, 2]
```

What about the total number of Waterfowl species?

```{r}
#| label: finding value for total number of waterfowl

group_species_count[4, 2]
```

Summarize these values in words.

In the 2025 water year, NCOS recorded:  

- 10 "species" of New World sparrows (9 species if considering Savannah Sparrow (Belding's) as a subset)
- 8 species of Shorebirds 
- 6 species of Herons, Ibis, and Allies 
- 13 species of Waterfowl 

## Total count per species

### Guiding question

What is the total count for each species in "Herons, Ibis, and Allies"/family Ardeidae for the 2025 water year?

### Pseudocode for summarizing

Write pseudocode to figure out the number of counted individuals in each species in this group.  

In no particular order, reference the following functions:  

- `arrange()` 
- `ungroup()`
- `filter()`
- `summarize()`
- `group_by()`

**Pseudocode:**

- create a new object from `birds_wy2025`
- filter to only include wading bird observations
- group by species
- calculate total counts per species
- ungroup the data frame
- arrange in descending order

### Running code

Then, write the code that corresponds to your pseudocode.

```{r}
#| label: calculating total count of herons

# create new object from birds_wy2025
herons_total_count <- birds_wy2025 |> 
  # filter to only include wading birds
  filter(lowest_group == "Ardeidae") |> 
  # group data frame by species
  group_by(species) |> 
  # calculate total counts per species
  summarize(total_count = sum(count)) |> 
  # ungroup data frame
  ungroup() |> 
  # arrange by descending order in total_count column
  arrange(-total_count)
```

### Visualizing total counts between species in Ardeidae

**Planning visualization:**  

- x-axis should be species
- y-axis should be total counts
- columns represent total counts in each species

#### Running code

```{r}
#| label: plotting total counts of herons

# base layer: ggplot
ggplot(data = herons_total_count,
       # x-axis should be species, y-axis should be count
       # fill columns by species
       mapping = aes(x = reorder(species, -total_count),
                     y = total_count,
                     fill = species)) +
  # first layer: column plot
  geom_col() +
  # take out legend
  theme(legend.position = "none") +
  scale_x_discrete(labels = label_wrap(15)) +
  # get rid of space between columns and x-axis
  scale_y_continuous(expand = c(0, 0),
                     limits = c(0, 42)) +
  # use custom colors
  scale_fill_manual(values = natparks.pals("Torres")) +
  # relabel x- and y-axes
  labs(x = "Species",
       y = "Total counts",
       title = "Herons, Ibis, and Allies (Ardeidae)") +
  # use different theme
  theme_bw() +
  # get rid of legend
  theme(legend.position = "none")
```

#### Interpretation

Black-crowned Night Herons have the highest total counts. White-faced Ibis have the lowest.

### What about Shorebirds?  

You can wrangle the data here by repurposing the code from above:

```{r}
#| label: calculating total count of shorebirds

# create new object from birds_wy2025
shorebirds_total_count <- birds_wy2025 |> 
  # filter to only include shorebirds
  filter(e_bird_group == "Shorebirds") |> 
  # group data frame by species
  group_by(species) |> 
  # calculate total counts per species
  summarize(total_count = sum(count)) |> 
  # ungroup data frame
  ungroup() |> 
  # arrange by descending order in total_count column
  arrange(-total_count)
```

And visualize:

```{r}
#| label: plotting total counts of shorebirds
#| fig-height: 6
#| fig-width: 8

# base layer: ggplot
ggplot(data = shorebirds_total_count,
       # x-axis should be species, y-axis should be count
       # fill columns by species
       mapping = aes(x = reorder(species, -total_count),
                     y = total_count,
                     fill = species)) +
  # first layer: column plot
  geom_col() +
  scale_x_discrete(labels = label_wrap(10)) +
  # get rid of space between columns and x-axis
  scale_y_continuous(expand = c(0, 0),
                     limits = c(0, 315)) +
  # use custom colors
  scale_fill_manual(values = natparks.pals("Olympic")) +
  # relabel x- and y-axes
  labs(x = "Species",
       y = "Total counts",
       title = "Shorebirds (Charadriiformes)") +
  # use different theme
  theme_bw() +
  # take out legend
  theme(legend.position = "none") 
```

#### Interpretation

Killdeer have the highest counts.  

Wilson's Snipe has the lowest.

## Seasonal differences in species richness

### Guiding question

In what season is species richness highest for each eBird group?

:::{.callout-note title="What is species richness?"}

Species richness is the _number of species that occur in a given sampling unit_. In this case, we are interested in the number of species in a given group in a given season in the 2025 water year. Thus, we are calculating _species richness_ for four different seasons for four different eBird groups.

:::

### Pseudocode for summarizing

When used with `group_by()` and `summarize()`, `n_distinct()` allows you to count the number of distinct values in a group or multiple groups.  

We can use this combination of functions to determine the number of species observed in a given eBird group in a given season.  

Write pseudocode to calculate the number of species observed in each eBird group in a given season. 

**Pseudocode:**

- create new object from `birds_wy2025`
- "roll up" Belding's subspecies to be considered with Savannah Sparrows
- group by ebird group and season
- count number of species in each group in each season 
- ungroup dataframe

:::{.callout-note title="What is roll up?"}
Belding's is a subspecies of Savannah Sparrow. When doing any calculations of species richness, it wouldn't make sense to include both Savannah Sparrow and Belding's Savannah Sparrow.

Instead of counting subspecies as species, researchers tend to "roll up" subspecies; that is, count subspecies as the higher species. In this example, we'll use `mutate()` and `case_when()` to roll up the subspecies to make sure that Belding's (the subspecies) is counted as Savannah (the species).
:::

### Running code

```{r}
#| label: calculating species richness per group per season

# create new object from birds_wy2025
season_spp_count <- birds_wy2025 |>
  # roll up subspecies
  mutate(species = case_when(
    # whenever Belding's appears, replace with Savannah Sparrow
    species == "Savannah Sparrow (Belding's)" ~ "Savannah Sparrow",
    # keep all other values in species
    .default = species
  )) |> 
  # group data frame by eBird group and season
  group_by(e_bird_group, season) |> 
  # count unique species that appear in each season in each group
  summarize(num_spp = n_distinct(species)) |> 
  # ungroup data frame
  ungroup() |> 
  # relevel season to be in correct order for water year
  mutate(season = fct_relevel(season,
                              "Fall", "Winter", "Spring", "Summer"))
```

### Visualizing

```{r}
#| label: plotting species richness per group per season

# base layer: ggplot
ggplot(data = season_spp_count,
       # x-axis should be season, y-axis should be number of species
       mapping = aes(x = season,
                     y = num_spp,
                     fill = season)) +
  # first layer: columns for species richness
  geom_col() +
  # custom color for seasons
  scale_fill_manual(values = c(
    "Fall" = "#287984",
    "Winter" = "#9c8020",
    "Spring" = "#589101",
    "Summer" = "#4987c9"
  )) +
  # get rid of space between columns and x-axis
  scale_y_continuous(expand = c(0, 0),
                     limits = c(0, 12)) +
  # different theme from default
  theme_bw() +
  # facet figure to have one panel for each eBird group
  facet_wrap(~ e_bird_group) +
  # relabel x-and y-axes
  labs(x = "Season",
       y = "Species richness") +
  # take out the legend
  theme(legend.position = "none")
  
```

### Interpretation

When is species richness highest and lowest for each group?

Species richness for Herons is highest in Spring and Summer, but lowest in Fall and Winter (low throughout though).  

Species richness for New World Sparrows is highest in Winter and lowest in summer.  

Shorebird species richness is highest in winter and spring, but lowest in fall and summer.  

Waterfowl species richness is highest in winter and spring and lowest in fall and summer. 

## Seasonal differences in total counts

### Guiding question

What are differences in total counts between seasons?  

How does this differ (or not) from differences in species richness between seasons?

:::{.callout-note title="What are total counts?"}

In each survey, some number of individuals are counted for a species.  

For example, you might see a flock of 12 Song Sparrows at the beginning of your survey, and another flock of 5 Song Sparrows at the end.  

For that survey, your total count of Song Sparrows is 17. However, there is no way to know for sure if the 5 sparrows from the end of your survey are different from the 12 you saw at the beginning.

For this reason, we can only calculate "total counts": the total number of counted individuals in a given survey.

:::

### Running code to summarize data

First, write code to calculate the number of individuals counted in each eBird group in each season. 

```{r}
#| label: calculating total counts per group per season

# create new object from birds_wy2025
season_bird_count <- birds_wy2025 |>
  # group data frame by eBird group and season
  group_by(e_bird_group, season) |> 
  # calculate total counts for each group in each season
  summarize(total_count = sum(count)) |> 
  # ungroup data frame
  ungroup() |> 
  # relevel seasons 
  mutate(season = fct_relevel(season,
                              "Fall", "Winter", "Spring", "Summer"))
```

### Visualizing data

Then, plot using `ggplot()`.

```{r}
#| label: plotting total count per group per season

# base layer: ggplot
ggplot(data = season_bird_count,
       # x-axis should be season, y-axis should be total counts
       # fill column by season
       mapping = aes(x = season,
                     y = total_count,
                     fill = season)) +
  # first layer: column plot
  geom_col() +
  # custom colors
  scale_fill_manual(values = c(
    "Fall" = "#287984",
    "Winter" = "#9c8020",
    "Spring" = "#589101",
    "Summer" = "#4987c9"
  )) +
  # get rid of space between bottom of column and x-axis
  scale_y_continuous(expand = c(0, 0),
                     limits = c(0, 400)) +
  # different theme from default
  theme_bw() +
  # facet so that there are four panels, one for each eBird group
  facet_wrap(~ e_bird_group) +
  # relabel x- and y-axes
  labs(x = "Season",
       y = "Total count") +
  # take out legend
  theme(legend.position = "none") 

```

### Interpretation

Are the patterns shown in this figure different from the patterns shown in the figure depicting species richness? Why or why not?

## Temporal dynamics of individual counts

### Guiding question

How do total counts change through time for each group?

### Summarize data

```{r}
#| label: calculating total counts per observation date per group

# create new object from birds_wy_2025
bird_count <- birds_wy2025 |> 
  # group data frame by observation date and eBird group
  group_by(observation_date, e_bird_group) |> 
  # calculate total counts for each date in each group
  summarize(total_count = sum(count)) |> 
  # ungroup data frame
  ungroup() 
```

### Visualize data

```{r}
#| label: plotting total counts per group

# base layer: ggplot
ggplot(data = bird_count,
       # x-axis should be observation date,
       # y-axis should be total_count,
       # color by group
       mapping = aes(x = observation_date,
                     y = total_count,
                     color = e_bird_group)) +
  # first layer: points
  geom_point() +
  # second layer: lines
  geom_line() +
  # custom colors
  scale_color_manual(values = natparks.pals("Saguaro")) +
  # different theme from default
  theme_bw() +
  # relabelling x- and y-axes
  labs(x = "Date",
       y = "Total count",
       color = "eBird group")
```

### Interpretation

Herons are consistently low.  

Waterfowl peak in winter and spring, then decline in summer.  

Shorebirds peak in fall and decline through winter, spring, and summer.

Sparrows peak in fall and decline through winter, spring, and summer.

## Temporal dynamics of species richness

### Guiding question

How does species richness change through time for each group?

### Summarizing the data

```{r}
#| label: calculating species richness for each survey

# creating new object from birds_wy2025
bird_spp_count <- birds_wy2025 |> 
  # roll up subspecies
  mutate(species = case_when(
    species == "Savannah Sparrow (Belding's)" ~ "Savannah Sparrow",
    .default = species
  )) |> 
  # group by observation date and eBird group
  group_by(observation_date, e_bird_group) |> 
  # calculate total counts
  summarize(sum_spp = n_distinct(species)) |> 
  # ungroup data frame
  ungroup()
```

### Visualizing the data

```{r}
#| label: visualizing species richness for each survey

# base layer: ggplot
ggplot(data = bird_spp_count,
       # x-axis should be observation date,
       # y-axis should be number of species
       # color by group
       mapping = aes(x = observation_date,
                     y = sum_spp,
                     color = e_bird_group)) +
  # first layer: points
  geom_point() +
  # second layer: lines
  geom_line() +
  # defining new y-axis breaks
  scale_y_continuous(breaks = seq(from = 0, to = 10, by = 2)) +
  # custom colors
  scale_color_manual(values = natparks.pals("Saguaro")) +
  # different theme from default
  theme_bw() +
  # relabelling axes and adding title
  labs(x = "Date",
       y = "Species richness",
       color = "eBird group")
```

### Interpretation

Species richness of Herons, Ibis, and Allies stays consistent throughout the year.

Species richness of waterfowl and shorebirds peaks in winter and spring.

Species richness of New World Sparrows is highest in fall/winter and decinds in spring and summer.

**END OF WEEK 2**










