---
title: "Water quality parameters"
subtitle: "Week 3"
categories: [library(), tidyverse, janitor, here, ggplot colors, filter(), mutate(), select(), case_when(), fct_relevel(), '%in%', group_by(), summarize(), ungroup(), count(), ggplot(), geom_point(), geom_line(), geom_bar(), facet_wrap(), scale_color_manual(), stat_summary(), mdy_hms(), month(), year(), left_join(), full_join(), pull(), '!', quarto]
---

[Download zip here](https://github.com/ENVS-193DD/ENVS-193DD.github.io/blob/main/data/week-03.zip)

# Summary

## Guiding questions

What are the differences in mean salinity between sites in the 2024 water year?

How do sites differ in salinity through time in the 2024 water year?

How does dissolved oxygen stratification differ through time at Venoco Bridge for the 2024 and 2025 water years?

## Data science tools

To answer these questions, we are going to:  

- review reading in packages
  - `library()`
- use familiar packages
  - `tidyverse`
- introduce new packages
  - `janitor` to clean column names
  - `here` to manage file paths
- review reading in data
  - `read_csv()`
- review cleaning and wrangling functions/operators
  - `filter()`
  - `mutate()`
  - `select()`
  - `case_when()`
  - `fct_relevel()`
  - `%in%`
- review summarizing functions 
  - `group_by()`, `summarize()`, and `ungroup()`
  - `count()`
- review visualizations using `ggplot()`, including:
  - `geom_point()` and `geom_line()`
  - `geom_bar()`
  - `facet_wrap()`
  - `scale_color_manual()`
- introduce a new visualization tool:
  - `stat_summary()`
- introduce new tools to deal with dates and times:
  - `mdy_hms()`
  - `month()`
  - `year()`
- introduce a new wrangling concept: **joins** using
  - `left_join()` and
  - `full_join()`
- introduce a new wrangling function:
  - `pull()`
- introduce a new operator:
  - `!` to exclude observations
- introduce new ways of formatting Quarto documents, including
  - [callout blocks](https://quarto.org/docs/authoring/callouts.html)
  - in-line code


# In-class code

## 1. Set up

### Packages

```{r}
#| label: packages
#| message: false

# general use
library(tidyverse)
library(janitor)

# file organization
library(here)
```

### Data

```{r}
#| label: data

# weather data from NOAA weather station
weather <- read_csv(here("data", "NOAA-weather-data.csv"))

# water quality survey metadata
metadata <- read_csv(here("data", "NCOS_YSI_Water_Quality_Monitoring_0.csv"))

# water quality parameter measurements
water_parameters <- read_csv(here("data", "YSI_Data_Begin_1.csv"))
```

### Understanding variables

#### Weather

- Which columns have the same value entered in each row?  

    - `STATION`
    - `NAME`
    - `LATITUDE`
    - `LONGITUDE`
    - `ELEVATION`

- Which columns have different values for each row?

    - `DATE`
    - `PRCP`
    - `TMAX`
    - `TMIN`

#### Metadata

- Observation location column names:

    - `Site Name`
    - `Other Site Name`
    - `x`
    - `y`

- Observation date and time column names:

    - `Monitoring Date`
    - `Time of WQ Measurement`

- Observation context column names:

    - `Weather`
    - `Flow`
    - `Notes`
    - `Water Surface Elevation`

- Person who took measurement:

    - `Monitor Name`

- ArcGIS Survey123 info:

    - `ObjectID`
    - `GlobalID`
    - `Creator`
    - `Editor`
    - `CreationDate`
    - `EditDate`

#### Water quality parameters

- ArcGIS Survey123 info column names:

    - `ObjectID`
    - `GlobalID`
    - `ParentGlobalID`
    - `Creator`
    - `Editor`
    - `CreationDate`
    - `EditDate`

- Measurement depth column names:

    - `Depth Code`

- Parameter column names:

    - `DO (mg/L)`
    - `DO (%sat)`
    - `Conductivity (uS/cm)`
    - `Salinity (ppt)`
    - `Temperature (°C)`

**`ParentGlobalID` in `water_parameters` is the same as `GlobalID` in `metadata`**

## 2. cleaning

### cleaning weather

#### Goal

Data frame with the following columns:  

- date
- precipitation, max temperature, min temperature
- month, year
- water year

#### Pseudocode

- create new object from weather
- clean column names
- make sure that date is understood as date
- select columns of interest
- extract month and year from date
- make new column for water year

#### Code

```{r}
#| label: cleaning-weather-data

# creating new object from weather
weather_clean <- weather |> 
  # cleaning column names
  clean_names() |> 
  # making sure date is in POSIXct format
  mutate(date = mdy(date)) |> 
  # selecting columns of interest
  select(date, prcp, tmax, tmin) |> 
  # extracting month and year from date column
  mutate(month = month(date),
         year = year(date)) |> 
  # assigning water year
  mutate(wy = case_when(
    # when month is October or later, assign value of year + 1
    month >= 10 ~ year + 1,
    # when month is September or earlier, assign value of existing year
    .default = year))
```

### cleaning metadata

#### Goal

Data frame with the following columns:

- ArcGIS Survey123 info:

    - global ID, 
    - creation date, 
    - edit date

- observation date and time:

    - monitoring date (note the format of this column), 
    - time of wq measurement
    
- observation context: 

    - weather, 
    - flow, 
    - notes, 
    - water surface elevation
  
- observation location: 

    - site name, 
    - other site name, 
    - x, 
    - y

#### Pseudocode

- create new object from metadata
- clean column names
- exclude unneeded columns
- make sure monitoring date is understood as date
- rename monitoring date column to show that it has datetimes
- extract month and year from monitoring date
- create new column for water year

#### Code

```{r}
#| label: cleaning-metadata

# creating new object from metadata
metadata_clean <- metadata |> 
  # cleaning column  names
  clean_names() |> 
  # excluding unnecessary columns
  select(!c(object_id, creator, editor)) |> 
  # making sure monitoring_date is in POSIXct format
  mutate(monitoring_date = mdy_hms(monitoring_date)) |> 
  # renaming monitoring datetime column to reflect values
  rename(monitoring_datetime = monitoring_date) |> 
  # creating new columns for month and year of observations
  mutate(month = month(monitoring_datetime),
         year = year(monitoring_datetime)) |> 
  # assigning water year
  mutate(wy = case_when(
    # when month is October or later, assign value of year + 1
    month >= 10 ~ year + 1,
    # when month is September or earlier, assign value of existing year
    .default = year)) |> 
  # create new column of full site name
  mutate(site_full = case_when(
    site_name == "east_channel" ~ "East Channel",
    site_name == "phelps_bridge" ~ "Phelps Bridge",
    site_name == "venoco_bridge" ~ "Venoco Bridge",
    site_name == "other" ~ "Other"
  ))
```

### cleaning water parameters

#### Goal

Data frame with the following columns:

- ArcGIS Survey123 info:

    - object ID, 
    - global ID, 
    - parent global ID, 
    - creation date, 
    - edit date

- measurement depth:

    - depth code
    
:::{.callout-note title="Consider the survey protocol"}
You can get information about the survey protocol from the Survey123 form.

In the "Depth code" field, the description is: Enter the correct code value for the depth you are monitoring, depending on the total depth of the site. For example, if the site si 2 feet deep, enter 2 for the surface, 1 for 1 foot depth, and 0 for the bottom. Bottom measurements will always be 0. If you are only able to take a surface measurement, enter 1.

Draw this out on a piece of paper. If a survey has depth codes of 0, 1, 2, 3, and 4, where would those fall relative to the bottom?
:::
    
- parameters:

    - DO (mg/L)
    - DO (%sat)
    - Conductivity (uS/cm)
    - Salinity (ppt)
    - Temperature (°C)
    
#### Pseudocode

- create new object from water parameters
- clean column names
- exclude unneeded columns
- rename depth code to be elevation 

#### Code

```{r}
#| label: cleaning-water-parameter-data

# creating new object from water_parameters
water_parameters_clean <- water_parameters |> 
  # cleaning object names
  clean_names() |> 
  # excluding unnecessary columns
  select(!c(creator, editor)) |> 
  # renaming columns
  rename(elevation_code = depth_code) |> 
  # replacing incorrect elevation code from 2021-07-16
  mutate(elevation_code = case_when(
    elevation_code == 10 ~ 1,
    .default = elevation_code
  )) |> 
  # set factors
  mutate(elevation_code = as_factor(elevation_code),
         elevation_code = fct_relevel(elevation_code,
                                      "0", "1", "2", "3", "4", "5", "6"))
```

## 3. exploratory visualization

### Temperature through time

#### Guiding question

What seasonal trends are there in maximum air temperature through time?

#### Planning

For a timeseries plot:  

- x-axis: `date`
- y-axis: `tmax`
- (optionally) filter to only include 2024 and 2025 water years

#### Code

```{r}
#| label: temp-timeseries
#| warning: false

# start with clean weather df
weather_clean |> 
  # filter to only include 2024 and 2025 water years
  filter(wy %in% c("2024", "2025")) |> 
  # x-axis: date, y-axis: maximum temp
  ggplot(mapping = aes(x = date,
                       y = tmax)) +
  # first layer: point geometry
  geom_point() +
  # second layer: line geometry
  geom_line()
```

#### Interpretation

**What seasonal trends do you see in maximum temperature?**

Air temperature is colder in winter months and hotter in summer months.

**What is the missing value, and with what date is it associated?**

There is a missing maximum temperature value from 2024-02-17.

### Number of observations at each elevation code

#### Planning

For bar plot:  

- x-axis: elevation code
- y-axis: counts

Necessary summarizing steps:

- group by elevation code
- count number of observations

#### Code

```{r}
#| label: elevation-counts

# create new object from clean water parameters df
elevation_obs <- water_parameters_clean |> 
  # group by elevation code
  group_by(elevation_code) |>
  # count number of observations at each depth
  count()

# start with elevation_obs df
ggplot(data = elevation_obs,
       # x-axis: elevation code
       # y-axis: count of observations
       mapping = aes(x = elevation_code,
                     y = n)) +
  # first layer: columns to represent counts
  geom_col()
```

#### Interpretation

**At which elevation code are there the most observations?**

Most observations are at elevation code 1, which is the surface.

**Are there any outliers in elevation codes?**

There is an elevation code of 10 from Venoco Bridge on July 16, 2021. Given that the water elevation was 4.5 ft, it is likely that this is supposed to be an elevation code of 1.  

To clean this up, we modified code in the `cleaning-water-parameters` chunk to replace elevation code 10 with 1. The new figure should not have an elevation code of 10.

### Number of observations in each water year

#### Planning

For bar plot:

- x-axis: water year
- y-axis: count
- facets: site name

Necessary wrangling:

- group by site name and water year
- count number of observations

#### Code

```{r}
#| label: site-counts

# creating new object from clean metadata
site_obs <- metadata_clean |> 
  # group by site and water year
  group_by(site_full, wy) |> 
  # count observations
  count()

# start with site_obs object
ggplot(data = site_obs,
       # x-axis: water year
       # y-axis: number of observations
       mapping = aes(x = wy,
                     y = n)) +
  # columns to represent count
  geom_col() +
  # facet to create panels for different sites
  facet_wrap(~ site_full,
             nrow = 1)
```

#### Interpretation

**How do sites differ in number of observations (across water years)?**

There are fewer observations at East Channel than at Phelps Bridge and Venoco Bridge, on average. This might be because of differences in water depth at each site.

**When do observations at "other" sites occur? Where are the "other" sites?**

:::{.callout-note title="Look at the `metadata_clean` object to help you answer this question!"}
Click on the `metadata_clean` object in the Environment. Search for "other" in the search bar.  

What sites are listed under `other_site_name`?
:::

These occur in the 2021 water year on June 25th and July 9th. The sites are Whittier Pond, Seasonal Pond, and Devereux Slough Pier.

There are three observations at "other" sites. 

## 4. Joining data frames

left join: keeps all observations in x

right join: keeps all observations in y

full join: keeps all observations in x and y

### water quality = metadata + water parameters

```{r}
#| label: joining-parameters-and-metadata

# join metadata and water parameters
water_quality <- full_join(
  # left
  x = metadata_clean, 
  # right
  y = water_parameters_clean,
  # shared column
  by = c("global_id" = "parent_global_id")) |> 
  # stop and look at colnames(water_quality) before continuing
  # selecting relevant columns
  select(
    # information about observation
    global_id, monitoring_datetime, time_of_wq_measurement, 
    monitor_name, weather, site_name, other_site_name, site_full, x, y, 
    
    # water information
    water_surface_elevation, flow, notes, 
    
    # measurement information
    elevation_code, do_mg_l, do_percent_sat, conductivity_u_s_cm,
    salinity_ppt, temperature_c) |> 
  # extract date from monitoring datetime
  mutate(date = date(monitoring_datetime)) |> 
  # move date column to appear to the right of monitoring_datetime
  relocate(date, .after = monitoring_datetime) |> 
  # exclude monitoring_datetime
  select(!monitoring_datetime)
```

### water quality weather = water quality + weather

```{r}
#| label: joining-water-quality-and-weather

# join water quality and weather
water_quality_weather <- full_join(
  # left
  x = water_quality, 
  # right
  y = weather_clean,
  # shared column
  by = "date"
)

# left join (exclude rows that don't match in weather)
water_quality_weather <- left_join(
  # left
  x = water_quality, 
  # right
  y = weather_clean,
  # shared column
  by = "date"
)
```

Where are the missing values? Which variables are missing values?

### Dealing with outliers

```{r}
#| label: salinity-outliers

# creating new object from water_quality_weather
salinity_outlier_df <- water_quality_weather |> 
  # filtering outliers
  filter(salinity_ppt > 1410) |> 
  # group by site 
  group_by(site_full) |> 
  # count number of outlier observations in each site
  count()

# creating new object
salinity_outlier_ids <- water_quality_weather |> 
  # filtering outliers
  filter(salinity_ppt > 1410) |> 
  # extracting global_id as vector for downstream use in filter
  pull(global_id)
```

There were `r salinity_outlier_df[2, 2]` outlier observations from Venoco Bridge, and `r salinity_outlier_df [1, 2]` outlier observation from East Channel (Dilling's Link). There are `r length(salinity_outlier_ids)` total outlier observations for salinity in this data set.

## 5. Visualizations

### Site differences in mean salinity

#### Guiding question

What are the differences in salinity between sites in the 2024 water year?

#### Planning

Goal: create a plot to compare differences in salinity between sites:

- x-axis should be site
- y-axis should be salinity
- points should represent individual salinity measurements at each site

Necessary wrangling steps:

- filter outliers
- exclude "other" sites
- filter to only include observations from 2024 water year
- filter to only include observations from bottom, ~1 ft above bottom, ~2 ft above bottom
- reorder site levels

#### Code

Wrangling:

```{r}
#| label: wrangling-salinity

# create new object from water_quality_weather
salinity_filtered <- water_quality_weather |> 
  # filter out outliers
  filter(!(global_id %in% salinity_outlier_ids)) |> 
  # filter out other sites
  filter(site_full != "Other") |> 
  # only include observations from 2024 water year
  filter(wy == "2024") |> 
  # filter bottom, ~ 1ft above bottom, ~2ft above bottom
  filter(elevation_code %in% c("0", "1", "2")) |> 
  # reorder site levels from lowest to highest mean salinity
  mutate(site_full = fct_relevel(
    site_full,
    "Phelps Bridge", "East Channel", "Venoco Bridge"
  ))
```

Quick summary:

```{r}
#| label: quick-summary

# start with filtered data frame
salinity_filtered |> 
  # group by site
  group_by(site_full) |> 
  # calculate mean salinity
  summarize(mean_salinity = mean(salinity_ppt, na.rm = TRUE)) |> 
  # ungroup data frame
  ungroup()
```

Visualizing:

```{r}
#| label: salinity-site-diff-plot

# base layer: ggplot
ggplot(data = salinity_filtered,
       # x-axis: site
       # y-axis: salinity
       mapping = aes(x = site_full,
                     y = salinity_ppt)) +
  # jitter to represent individual salinity measurements
  geom_jitter(height = 0,
              width = 0.2,
              shape = 21) +
  # quick calculation of mean for visualization
  stat_summary(fun = mean,
               geom = "point",
               color = "red",
               size = 5) +
  # relabelling x- and y-axes, adding title
  labs(x = "Site",
       y = "Salinity (ppt)",
       title = "Sites differ in mean salinity across all timepoints") +
  # different theme from default
  theme_bw()
```


#### Interpretation

**How do sites differ in mean salinity across all timepoints?**

### Temporal salinity dynamics

#### Guiding question

How do sites differ in salinity through time in the 2024 water year?

#### Planning

Goal: create a figure with:

- time on the x-axis
- salinity on the y-axis
- different colors for each elevation code
- different panels for each site

No wrangling needed if using `salinity_filtered` data frame!

#### Code

Visualizing:

```{r}
#| label: salinity-through-time-by-site
#| fig-height: 6
#| fig-width: 24

# using salinity_filtered data frame
ggplot(data = salinity_filtered,
       # x-axis: date
       # y-axis: salinity
       # color by elevation code
       mapping = aes(x = date,
                     y = salinity_ppt,
                     color = elevation_code)) +
  # points to represent measurements 
  geom_point() +
  # lines to connect measurements
  geom_line() +
  # separate panels for each site
  facet_wrap(~ site_full) +
  # custom colors
  scale_color_manual(
    values = c("0" = "darkgreen",
               "1" = "goldenrod",
               "2" = "turquoise")
  ) +
  # relabelling x-and y-axes, legend
  labs(x = "Date", 
       y = "Salinity (ppt)",
       color = "Elevation code") +
  # different theme
  theme_bw()
```

#### Interpretation

**Are there differences in trends through time between sites?**

**Are there differences in elevation through time between sites?**

### Dissolved oxygen stratification

#### Guiding question

How does dissolved oxygen stratification differ through time at Venoco Bridge for the 2024 and 2025 water years?

#### Planning

Goal: create a figure with:

- time on the x-axis
- percent saturation of DO on y-axis
- colors should represent elevation codes
- different panels for water years

Necessary cleaning steps:

- filter to only include Venoco, water years, elevation codes
- select columns of interest

#### Code

Wrangling:

```{r}
#| label: do-dataframe

# start with joined data frame
venoco_do <- water_quality_weather |> 
  # filter to only include Venoco,
  filter(site_full == "Venoco Bridge" &
           # water years of interest
           wy %in% c("2024", "2025") &
           # bottom, ~ 1 ft, ~ 2 ft above bottom
           elevation_code %in% c("0", "1", "2")) |> 
  # select columns of interest
  select(date, wy, do_mg_l, elevation_code) |> 
  # relevel elevation code order
  mutate(elevation_code = fct_relevel(elevation_code,
                                      "2", "1", "0"))
```

Visualizing:

```{r}
#| label: plot-DO
#| fig-height: 4
#| fig-width: 14

# start with venoco DO data frame
ggplot(data = venoco_do,
       # x-axis: date
       # y-axis: percent saturation dissolved oxygen
       # color by elevation code
       mapping = aes(x = date,
                     y = do_mg_l,
                     color = elevation_code)) +
  # points to represent DO measurements
  geom_point() +
  # lines to connect points
  geom_line() +
  # separate panels for each water year
  # free x-axis 
  facet_wrap(~ wy,
             scales = "free_x") +
  # custom colors
  scale_color_manual(
    values = c("2" = "turquoise", 
               "1" = "goldenrod", 
               "0" = "darkgreen")
  ) +
  # different theme
  theme_bw() +
  # relabelling
  labs(x = "Date",
       y = "Dissolved oxygen (mg/L)",
       color = "Elevation code")
```


#### Interpretation

**Do water years differ in the degree of stratification in winter months?**

Stratification is more obvious in the 2025 water year than in the 2024 water year.

**Does stratification change through time?**

The magnitude of stratification is greater in winter than in spring and summer, though there is some stratification in summer even though DO is lower overall.
